Ecological performance of detecting data fabrication

### 2017-07-03 submission/manuscript.Rmd
Revisions intro, add results study 2

---
title: "Ecological performance of detecting data fabrication"
author: "Chris HJ Hartgerink, Jan G Voelkel, Jelte M Wicherts, Marcel ALM van Assen"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  word_document: default
  html_document:
    toc: yes
csl: ../bibliography/apa.csl
bibliography: ../bibliography/library2.bib
---

```{r preparation, echo = FALSE}
suppressPackageStartupMessages(if(!require(pROC)){install.packages('pROC', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(pROC))
suppressPackageStartupMessages(if(!require(foreign)){install.packages('foreign', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(foreign))
suppressPackageStartupMessages(if(!require(latex2exp)){install.packages('latex2exp', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(latex2exp))
suppressPackageStartupMessages(if(!require(ggplot2)){install.packages('ggplot2', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(if(!require(gridExtra)){install.packages('gridExtra', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(if(!require(xtable)){install.packages('xtable', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(xtable))
suppressPackageStartupMessages(if(!require(httr)){install.packages('httr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(httr))
suppressPackageStartupMessages(if (!require(car)){install.packages('car', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(car))
suppressPackageStartupMessages(if (!require(effects)){install.packages('effects', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(effects))
suppressPackageStartupMessages(if (!require(lsr)){install.packages('lsr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(lsr))
suppressPackageStartupMessages(if (!require(data.table)){install.packages('data.table', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(if (!require(dplyr)){install.packages('dplyr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(if (!require(reshape2)){install.packages('reshape2', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(reshape2))

@@

source(sprintf('%sfunctions/digit_analysis.R', x12))
source(sprintf('%sfunctions/std_var.R', x12))

@@

Any field of empirical inquiry is faced with cases of scientific misconduct at some point, either in the form of fabrication, falsification or plagiarism (FFP). Psychology was faced with Stapel; medical sciences were faced with Poldermans and Macchiarini; life sciences were faced with Voignet. These are just a few examples of misconduct cases in the last decade. Overall, an estimated 2\% of all scholars have admitted to falsifying or fabricating research results at least once [@10.1371/journal.pone.0005738] and this is likely to be an underestimate due to socially desirable responses. The detection rate is likely to be even lower; for example, only around a dozen cases are discovered in the United States and the Netherlands, despite covering several hundreds of thousands of researchers. At best, this amounts to a detection rate far below 1\% of those 2\% who admit to fabricating data --- the tip of a seemingly much larger iceberg.

In order to stifle attempts at data fabrication, improved detection of fabricated data is considered to deter such harmful attempts. 
Although deterrence theory dates back to the middle of the 17th century [@leviathan], its implementation has not occurred across the different forms of scientific misconduct equilaterally. Basically, deterrence theory stipulates that with increased risk of detection, the utility of scientific misconduct (for this context) will decrease and therefore fewer people will engage in such behaviors. This principle of deterrence has been implemented with plagiarism scanners, a development that already started a long time ago [e.g., @10.1109/13.28038]. However, increased deterrence of fabrication and falsification by improved detection mechanisms has not been as widely implemented.

In the last decade, detecting image manipulation has become one of the few forms of detecting scientific misconduct (alongside plagiarism).
The Journal of Cell Biology scans each submitted image for potential manipulation [@The_Journal_of_Cell_Biology2015-vh], which greatly increases the risk of detecting (blatant) image manipulation. More recently, algorithms have been developed to automate the scanning of images for (subtle) manipulations [@10.1007/s11948-016-9841-7]. These developments in detecting image manipulation have increased detection risk during the pre-publication and post-publication phase by improving detection mechanisms and increasing the understanding of how images might be manipulated. Moreover, their application also helps researchers systematically evaluate research articles to estimate the extent of the problem of image manipulation [4\% of all papers are estimated to contain manipulated images, @10.1128/mBio.00809-16].

Statistical methods can provide one way to improve detection of data fabrication in empirical research. Humans are notoriously bad at understanding and estimating probabilities [e.g., @10.1126/science.185.4157.1124;@10.1037/h0031322], which could manifest itself in the fundamentally probabilistic data they try to fabricate. 
That researchers do not understand probabilistic processes also presents itself in false interpretation of genuine research data [@10.3758/BF03213921;@10.1007/s11336-015-9446-0;@10.7326/0003-4819-130-12-199906150-00008]. When data are fabricated, probabilistic principles are easily violated if forgotten at the univariate level, bivariate level, trivatiate level, or beyond [@Haldane1948-nm]. Based on this idea, statistical methods that investigate whether the reported data are feasible under the theoretically probabilistic processes can be used to detect potential data fabrication.

The application of such statistical methods to detect data fabrication has occurred in several cases in recent years and has potential for future application beyond a case-basis. 
For example, problems in papers by Fuji were highlighted with statistical methods [@10.1111/j.1365-2044.2012.07128.x;@10.1111/anae.13126], resulting in 183 retractions [@oransky2015]. In this case, baseline measures across randomized groups were examined for too little variation. Random assignment should introduce a certain degree of random error that might be missed by a human fabricator, misestimating the probabilistic process that generates such error. Another two cases are those of Sanna and Smeesters, where fabricated data were also detected with statistics [@10.1177/0956797613480366]. These cases inspected the variance of variances (i.e., the second, or meta level variance). Once again, too little variation was what revealed problems in these data [other examples include @lacour-complaint;@foerster-complaint]. These methods, although developed in a case-setting, need not be limited to cases. The application of such methods can be (semi-)automated if data are available in a machine-readable format that one of the statistical methods can be applied to. An example of such a potential case for mass application of using statistics to detect (potential) data fabrication is in the ClinicalTrials.gov database, where baseline measures across randomized groups are readily available for download and subsequent analysis [@10.3897/rio.1.e7462].

Nonetheless, considering the potential harm of applying statistical methods to flag potentially problematic results, it needs to be sorted out whether such methods have diagnosticity that actually makes it responsible to apply them. 
We hardly know how researchers might go about fabricating data. Cases such as Fuji, Smeesters, and Sanna provide some insights, but are highly pre-selected (i.e., those who got caught/confessed) and as such, systematically biased. Relatively extensive descriptions in rare and partial autobiogrophical accounts provide little insight into the actual data fabrication process, except for the setting where it might take place [e.g., late at night when no one is around; @stapel_book]. However, trustworthiness of these accounts can be called into question. Additionally, the performance of methods to detect data fabrication is highly dependent on the unknown prevalence of data fabrication and the power to actually to detect data fabrication. Given that we do not know how researchers might fabricate data, the diagnosticity of these methods cannot realistically be simulated.

The effectiveness of detection mechanisms and their consequences, hence their expected deterrence, is exacerbated in the digital age by the increased usage of public online discussion platforms such as [PubPeer (https://pubpeer.com)](https://pubpeer.com). 
PubPeer serves as an "online journal club" where anyone can discuss articles. Authors are notified when someone leaves a response, providing them with the possibility to respond. Such a platform allows for public discussion of the paper, including discussion of reanalyses, methodology, availability of materials, etc. When detection mechanisms are freely available to use, they can lead to (a surge of) comments when applied by users. Recently, in 2016, one user (the main author of this paper) used `statcheck` software to report potential statistical reporting inconsistencies for 50,000 psychology articles, which led to a large discussion about (automated) online comments [@10.1038/540151a]. The impact of such new possibilities is not to be underestimated, although its potential to constructively contribute to the scientific discussion should also not be [@10.1038/540007b]. Nonetheless, the `statcheck` software was well validated prior to this application [@10.3758/s13428-015-0664-2] and the same principle applies to the application of statistical methods to detect (potential) data fabrication.

@@

Throughout this paper, we inspect statistical methods to detect data fabrication that can be applied to (1) summary results or (2) raw data. 
Even though structure and contents of data can look different depending on the structure of a study and the measures, there are certain common characteristics of empirical results and the underlying raw data that can be inspected. For example, summary results frequently include means, standard deviations, test-statistics, and $p$-values. Raw data frequently contain at least some variables measured at a interval- or ratio scale [@10.1126/science.103.2684.677]. Such common characteristics allow for the development of generic statistical methods that can be applied across a varied set of results. We review the theoretical framework of the specific methods we apply throughout this paper, but these are not exhaustive of all methods available to test for potential problems in empirical data [see also @buyse1999;@10.7287/peerj.preprints.2400v1;@10.7287/peerj.preprints.2064v1;@sprite-heathers].

@@

The distribution of one $p$-value depends on various parameters, but there are specific theoretical boundary conditions to this distribution [@fisher1925]. Based on the population effect size, the precision of the estimate, and the observed effect size, the $p$-value distribution can be uniform (when then null hypothesis is true) or right-skewed (when the alternative hypothesis is true). By extension, any other distribution, after taking into account sampling error, is theoretically suspect. For example, a distribution might show a bump when authors $p$-hack [e.g., Figure 1 in @10.7717/peerj.1935], but fabricators might create other non-uniform or non-right-skewed distributions, failing to take into account the theoretical boundary conditions.

In order to test whether observed $p$-values transgress the boundary condition, we proposed an adaptation of Fisher's method [@10.1186/s41073-016-0012-9]. This adaptation is a simple reversal of the Fisher method [@fisher1925], which was originally introduced as a simple meta-analytic test to indicate presence of an effect; this test is computed as 
$$
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(p_i)
$$
where it tests for right-skew (i.e., more smaller $p$-values than larger $p$-values) across the $k$ number of $p$-values. Reversing the Fisher method results in 
$$
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(1-\frac{p_i-t}{1-t})
$$
where it now tests for left-skew (i.e., more larger $p$-values than smaller $p$-values) across the $k$ number of $p$-values that falls above the threshold $t$. This threshold is added and can be set to anything. Given the frequent misinterpretation of $p$-values [e.g., as the probability of an effect, @10.1053/j.seminhematol.2008.04.003;@10.1136/bmj.311.7003.485], researchers who fabricate nonsignificant data might forget to fabricate a uniform $p$-value distribution and generate only large $p$-values when fabricating nonsignificant results. In that case, it makes  sense to set $t=.05$ or $t=.10$, in order to include only nonsignificant test results.

Despite this test being useful for detecting data anomalies in nonsignificant $p$-values, two clear exceptions should be taken into account: (1) model-fit tests and (2) wrongly specified one-tailed tests. When model-fit tests are used, these can quickly result in high $p$-values because of model saturation (i.e., where the null hypothesis is good fit and the alternative is bad fit). For properly specified one-tailed tests, the $p$-value distribution is right-skewed. When wrongly specified, this distribution is reversed and becomes left-skew. These two exceptional cases should not be forgotten during the application of this test, but are irrelevant for the rest of this paper due to our study design (see methods of both studies).

@@

Variance- or standard deviation estimates are typically reported to indicate spread, but just like the mean there should be sampling error in this estimate (i.e., the variance of the variances) proportionate to the sample size. A variance estimate follows a $\chi^2$-distribution, which is dependent on the sample size [p. 445; @hogg-tanis]; that is
$$
z^2_j\sim\left(\frac{\chi^2_{N_j-1}}{N_j-1}\right)/MS_w
$$
where $N_j$ is the sample size of the $j$th group and $MS_w$ is the normalizing constant resulting in a standardized variance $z^2_j$. The normalizing constant $MS_w$ is computed as 
$$
MS_w=\frac{\sum\limits^k_{j=1}(N_j-1)s^2_j}{\sum\limits^k_{j=1}(N_j-1)}
$$
where $s^2_j$ is the variance in the $j$th group. If simulating (see next paragraph), the simulated variances are used to compute $MS_w$.

Assuming that the observed variances are from the same population distribution, the expected spread of reported standardized variances can be simulated. By repeatedly drawing (standardized) variance estimates for $j$ groups and then computing their spread allows for approximation of the sampling distribution. Spread can be operationalized in various ways, such as the standard deviation of the variances (denoted in this paper as $SD_z$) or as the range of the variances (denoted as $max-min_z$). If there is reason to believe that the observed variances come from heterogeneous populations, subgrouping (i.e., one variance analysis per subgroup) is necessary to prevent false results.

Subsequently, the observed sampling variance (or range) can be compared to the expected sampling distribution. Too consistent results would indicate potential anomalies in the reported data [@10.1177/0956797613480366]. For example, if four independent samples all yield the variance $2.22$, this could be considered excessively consistent when the probability that this amount of consistency (or more) is less than 1 out of 1000 in truly random samples.

@@

A fabricator might present unrealistically large effects, forgetting the exponential relation between the effect size measure and the implied correlation. From our own experience, and anecdotal evidence elsewhere [@10.1016/0197-24569190037-M], large effects have previously raised initial suspicions. The size of a large effect can quickly become unrealistic; $d=1.2$ already implies a correlation of `r round(sqrt((1.2^2/(1.2^2+4))), 2)` between the dependent- and independent measures. More generally, the relation between the implied correlation and the effect size tends to be exponential, as Figure `r fignr` depicts.

```{r d2, fig.cap="_Figure XX._ The relation between Cohen's d effect size and its direct transformation, the correlation r (or vice versa). If effects d>2 are fabricated, the implied correlation between the dependent- and independent measure(s) is approximately .6.", out.width='100%', echo=FALSE}
d2 <- seq(0, 10, .01)

plot(sqrt(d2^2/(d2^2+4)), x = d2,
     type='l', xlab = "Cohen's d", ylab = "Correlation r")

Taking the observed effect size and transforming it into a correlation, allows for an easy way to assess how extreme the presented result is. One minus the observed correlation can be used as a measure for extreme effects (i.e., $1-r$); as a heuristic, it can be regarded as a $p$-value. That is, this measure too ranges from zero to one and the more extreme the effect size, the smaller the value. This method specifically looks at situations where fabricators would want to fabricate the existence of an effect (not the absence of one).

@@

>>>>

---
title: "Ecological performance of detecting data fabrication"
author: "Chris HJ Hartgerink, Jan G Voelkel, Jelte M Wicherts, Marcel ALM van Assen"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_depth: 2
csl: ../bibliography/apa.csl
bibliography: ../bibliography/library2.bib
---

```{r preparation, echo = FALSE}
suppressPackageStartupMessages(if(!require(pROC)){install.packages('pROC', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(pROC))
suppressPackageStartupMessages(if(!require(knitr)){install.packages('knitr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(if(!require(foreign)){install.packages('foreign', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(foreign))
suppressPackageStartupMessages(if(!require(latex2exp)){install.packages('latex2exp', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(latex2exp))
suppressPackageStartupMessages(if(!require(ggplot2)){install.packages('ggplot2', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(if(!require(gridExtra)){install.packages('gridExtra', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(if(!require(xtable)){install.packages('xtable', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(xtable))
suppressPackageStartupMessages(if(!require(httr)){install.packages('httr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(httr))
suppressPackageStartupMessages(if (!require(car)){install.packages('car', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(car))
suppressPackageStartupMessages(if (!require(effects)){install.packages('effects', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(effects))
suppressPackageStartupMessages(if (!require(lsr)){install.packages('lsr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(lsr))
suppressPackageStartupMessages(if (!require(data.table)){install.packages('data.table', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(if (!require(dplyr)){install.packages('dplyr', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(if (!require(reshape2)){install.packages('reshape2', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(if (!require(metafor)){install.packages('metafor', repos = 'http://cran.us.r-project.org')})
suppressPackageStartupMessages(library(metafor))

@@

source(sprintf('%sfunctions/digit_analysis.R', x12))
source(sprintf('%sfunctions/std_var.R', x12))
source(sprintf('%sfunctions/fisher_method.R', x12))

@@

Any field of empirical inquiry is faced with cases of scientific misconduct at some point, either in the form of fabrication, falsification or plagiarism (FFP). Psychology was faced with Stapel; medical sciences were faced with Poldermans and Macchiarini; life sciences were faced with Voignet. These are just a few examples of misconduct cases in the last decade. Overall, an estimated 2\% of all scholars have admitted to falsifying or fabricating research results at least once [@10.1371/journal.pone.0005738] and this is likely to be an underestimate due to socially desirable responses. The detection rate of data fabrication is likely to be even lower; for example, only around a dozen cases are discovered in the United States and the Netherlands, despite covering several hundreds of thousands of researchers. At best, this amounts to a detection rate far below 1\% of those 2\% who admit to fabricating data --- the tip of a seemingly much larger iceberg.

In order to stifle attempts at data fabrication, improved detection of fabricated data is considered to deter such harmful attempts. 
Although deterrence theory dates back to the middle of the 17th century [@leviathan], its implementation has not occurred equally across fabrication, falsification, and plagiarism. Basically, deterrence theory stipulates that with increased risk of detection, the utility of scientific misconduct (for this context) will decrease and therefore fewer people will engage in such behaviors. This principle of deterrence has been implemented with plagiarism scanners, a development that already started a long time ago [e.g., @10.1109/13.28038]. In the last decade, detecting image manipulation has become one of the few forms of detecting data fabrication . The Journal of Cell Biology and EMBO scan each submitted image for potential manipulation [@The_Journal_of_Cell_Biology2015-vh;@10.1038/546575a], which greatly increases the risk of detecting (blatant) image manipulation. More recently, algorithms have been developed to automate the scanning of images for (subtle) manipulations [@10.1007/s11948-016-9841-7]. These developments in detecting image manipulation have increased detection risk during the pre-publication and post-publication phase by improving detection mechanisms and increasing the understanding of how images might be manipulated. Moreover, their application also helps researchers systematically evaluate research articles to estimate the extent of the problem of image manipulation [4\% of all papers are estimated to contain manipulated images, @10.1128/mBio.00809-16].

Statistical methods provide one way to improve detection of data fabrication in empirical research. Humans are notoriously bad at understanding and estimating probabilities [e.g., @10.1126/science.185.4157.1124;@10.1037/h0031322], which could manifest itself in the fundamentally probabilistic data they try to fabricate. 
When data are fabricated, probabilistic principles are easily violated if forgotten at the univariate level, bivariate level, trivatiate level, or beyond [@Haldane1948-nm]. Based on this idea, statistical methods that investigate whether the reported data are feasible under the theoretically probabilistic processes can be used to detect potential data fabrication.

```{r echo=FALSE}
set.seed(1234)
ex1 <- t(replicate(n = 10, expr = table(sample(x = c("M", "F"), size = 100, replace = TRUE))))
ex1p <- pchisq(apply(ex1, 1, function(x) sum((x-50)^2/50)), df = 1, lower.tail = FALSE)
set.seed(1234)
ex2_1 <- 50 + sample(x = -1:1, size = 10, replace = TRUE)
ex2_2 <- 100 - ex2_1
ex2 <- cbind(ex2_1, ex2_2)
ex2p <- pchisq(apply(ex2, 1, function(x) sum((x-50)^2/50)), df = 1, lower.tail = FALSE)

df <- data.frame(study = sprintf("Study %s", 1:10),
                 ex1,
                 ex1p,
                 ex2,
                 ex2p)
names(df) <- c("Study",
               "Experimental [S1]",
               "Control [S1]",
               "P-value [S1]",
               "Experimental [S2]",
               "Control [S2]",
               "P-value [S2]")
```

The application of such statistical methods to detect data fabrication has occurred in several cases of scientific misconduct in recent years and has potential for future application beyond these specific cases. 
For example, problems in papers by Fuji were highlighted with statistical methods [@10.1111/j.1365-2044.2012.07128.x;@10.1111/anae.13126], resulting in 183 retractions [@oransky2015]. This method works as follows: in true randomized clinical trials (RCTs), baseline measurements should be statistically equal across groups. As such, the $p$-values for group comparisons would be expected to be uniformly distributed because the null hypothesis of equal groups is true by definition of the design. In the Fuji papers, group comparisons showed excessive consistency, resulting primarily in high $p$-values (e.g., .99, .95) and a high mean $p$-value across the comparisons, where a mean $p$-value of .5 is expected. As an illustration, see Table `r tabnr`, which depicts the assignment of participants to the experimental or control condition for a true randomized design (Set 1) or for a fabricated design (Set 2). The mean $p$-value for the true randomized design Set 1 is `r round(mean(ex1p), 3)`, whereas the fabricated Set 2 has mean $p$-value `r round(mean(ex2p), 3)`. Other statistical methods to detect data fabrication are addressed in the theoretical framework.
.

```{r table_excess, echo = FALSE}
knitr::kable(df, digits = 3, caption = "Examples of participant counts for genuine- and fabricated randomized clinical trials. Set 1 (S1) is randomly generated data under the null hypothesis of random assignment (assumed to be the genuine process), whereas Set 2 (S2) is generated under excessive consistency with equal groups. Each trial contains 100 participants. The $p$-values ")

tabnr <- tabnr + 1
```

Such statistical methods to detect data anomalies, although developed to quantify suspicions in a specific paper, could be applied to screen multiple papers. The application of such methods can be (semi-)automated if data are available in a machine-readable format that one of the statistical methods can be applied to. An example of such a potential case for mass application of using statistics to detect (potential) data fabrication is in the ClinicalTrials.gov database, where baseline measures across randomized groups are readily available for download and subsequent analysis [@10.3897/rio.1.e7462].

Nonetheless, prior to applying statistical methods to flag potentially problematic results, investigating whether such methods function well enough in detecting problems is required for responsible application. 
We hardly know how researchers might go about fabricating data. Cases such as Stapel, Fuji, Smeesters, and Sanna provide some insights, but are highly pre-selected (i.e., those who got caught/confessed) and as such, systematically biased. Relatively extensive descriptions in rare and partial autobiogrophical accounts provide little insight into the actual data fabrication process, except for the setting where it might take place [e.g., late at night when no one is around; @stapel_book]. However, trustworthiness of these accounts can be called into question. Additionally, the performance of methods to detect data fabrication is highly dependent on the unknown prevalence of data fabrication and the power to actually to detect data fabrication. Given that we do not know how researchers might fabricate data, the diagnosticity of these methods cannot realistically be assessed.

Throughout this paper, we inspect statistical methods to detect potential data fabrication that can be applied to (1) summary results or (2) raw data. 
Even though structure and contents of data can look different depending on the structure of a study and the measures, there are certain common characteristics of empirical results and the underlying raw data that can be inspected. For example, summary results frequently include means, standard deviations, test-statistics, and $p$-values. Raw data frequently contain at least some variables measured at a interval- or ratio scale [@10.1126/science.103.2684.677]. Such common characteristics allow for the development of generic statistical methods that can be applied across a varied set of results to screen for problematic data. We review the theoretical framework of the specific methods we apply throughout this paper, but these are not exhaustive of all methods available to test for potential problems in empirical data [see also @buyse1999;@10.7287/peerj.preprints.2400v1;@10.7287/peerj.preprints.2064v1;@sprite-heathers].

@@

In the current paper, we differentiate between statistical methods to detect potential data fabrication based on reported data (i.e., summary results) or raw data. For summary results, we review $p$-value analysis, variance analysis, and effect size analysis as potential ways to detect data anomalies. $P$-value analysis can be applied whenever sufficient nonsignificant $p$-values are reported; variance analysis can be applied whenever a set of variances are reported for independent groups alongside the sample sizes per group; effect size analysis can be used whenever the effect size can be computed [e.g., an APA reported $t$- or $F$-statistic;@10.1525/collabra.71].  For raw data, we review digit analyses (i.e., the Newcomb-Benford law and terminal digit analysis), and multivariate associations as potential ways to detect data anomalies. Digit analyses can be applied when ratio scale measures are present in the raw data; multivariate associations can be applied whenever there is a multivariate relation in the raw data that can be computed and control data is available from the literature.

@@

The distribution of a $p$-value is uniform if the null hypothesis is true and right-skewed if the alternative hypothesis is true [@fisher1925]. One $p$-value is the result of the population effect size, the precision of the estimate, and the observed effect size, whose properties carry over to a set of $p$-values if those $p$-values are independent. As such, the $p$-value distribution of a set of $p$-values is uniform when the null hypothesis is true, or right-skewed when the alternative hypothesis is true.

When the $p$-value distribution is not uniform nor right-skewed, it can indicate potential data anomalies. For example, a distribution might show a bump when excessive $p$-hacking occurs that seeks barely significant results [e.g., Figure 1 in @10.7717/peerj.1935]. However, data fabricators might create other non-uniform or non-right-skewed distributions, failing to take into account the theoretical boundary conditions. For example, given the frequent misinterpretation of $p$-values [@10.3758/BF03213921;@10.1007/s11336-015-9446-0;@10.7326/0003-4819-130-12-199906150-00008], it is possible that the same underlying misinterpration will manifest itself in fabricated data. More specifically, given the frequent misinterpretation of $p$-valuesas the probability of an effect [@10.1053/j.seminhematol.2008.04.003;@10.1136/bmj.311.7003.485], researchers who fabricate nonsignificant data might forget to fabricate a uniform $p$-value distribution and generate only large $p$-values when fabricating nonsignificant results.

In order to test whether observed $p$-values are anomalous, we proposed an adaptation of Fisher's method [@10.1186/s41073-016-0012-9]. This adaptation is a simple reversal of the original Fisher method [@fisher1925], which was introduced as a meta-analytic test for the presence of an effect. This test is computed as 
$$
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(p_i)
$$
where it tests for right-skew (i.e., more smaller $p$-values than larger $p$-values) across the $k$ number of $p$-values. Reversing the Fisher method results in 
$$
\chi^2_{2k}=-2\sum\limits^k_{i=1}\ln(1-\frac{p_i-t}{1-t})
$$
where it now tests for left-skew (i.e., more larger $p$-values than smaller $p$-values) across the $k$ number of $p$-values that falls above the threshold $t$. This threshold is added and can be set to any value between zero and one. Upon writing this paper, it became clear to us that this is reversed Fisher method is similar to the operating principle of Carlisle's method testing for excessive homogeneity across baseline measurements in RCTs [@10.1111/anae.13938;@10.1111/j.1365-2044.2012.07128.x;@10.1111/anae.13126].

Despite this test being useful for detecting data anomalies in nonsignificant $p$-values, one exception should be taken into account: wrongly specified one-tailed tests. For properly specified one-tailed tests, the $p$-value distribution is right-skewed. When wrongly specified, this distribution is reversed and becomes left-skew. As such, any data anomalies detected with this method would need to be inspected for misspecified one-tailed hypotheses to preclude false conclusions.

@@

Variance- or standard deviation estimates are typically reported to indicate dispersion, but just like the mean there should be sampling error in this estimate proportional to the sample size [i.e., $\sigma/\sqrt{2n}$ under the assumption of normality, p. 351;@yule1922]. A variance estimate follows a $\chi^2$-distribution, which is dependent on the sample size [p. 445; @hogg-tanis]; that is
$$
z^2_j\sim\left(\frac{\chi^2_{N_j-1}}{N_j-1}\right)/MS_w
$$
where $N_j$ is the sample size of the $j$th group and $MS_w$ is the normalizing constant resulting in a standardized variance $z^2_j$. The normalizing constant $MS_w$ is computed as 
$$
MS_w=\frac{\sum\limits^k_{j=1}(N_j-1)s^2_j}{\sum\limits^k_{j=1}(N_j-1)}
$$
where $s^2_j$ is the variance in the $j$th group.

The observed dispersion of the variances can be compared to various measures of expected dispersion in the variances. Dispersion can be operationalized in various ways, such as the standard deviation of the variances [denoted in this paper as $SD_z$;@10.1177/0956797613480366] or as the range of the variances (denoted as $max-min_z$). Too consistent results would indicate potential anomalies in the reported data. For example, in the Smeesters case three independent conditions from the same study ($n_k=15$) were reported to have standard deviations 25.09, 24.58, and 25.65. The standard deviation of the standard deviations here is `r round(sd(c(25.09, 24.58, 25.65)), 2)` (i.e., $SD_z$). Such consistency (or more consistency) would only be observed in `r round(std_var(n = rep(15, 3), sds = c(25.09, 24.58, 25.65), iter = 100000)*100, 2)`% of 100,000 simulated replications [@10.1177/0956797613480366].

@@

From our own experience, and anecdotal evidence elsewhere [@10.1016/0197-24569190037-M], large effects have previously raised initial suspicions. Taking the observed effect size and transforming it into a correlation, allows for an easy way to assess how extreme the presented result is. One minus the observed correlation can be used as a measure for extreme effects (i.e., $1-r$); as a heuristic, it can be regarded as a $p$-value. That is, this measure too ranges from zero to one and the more extreme the effect size, the smaller the value. This method specifically looks at situations where fabricators would want to fabricate the existence of an effect (not the absence of one).

@@

Raw data with ratio- or interval measures can be subjected to digit analysis under specific conditions. More specifically, the properties of the leading (first) digit (e.g., the 1 in 123.45) or the terminal (last) digit (e.g., the 5 in 123.45) can be examined. By analyzing these leading- and terminal digits for deviations from specific digit distributions, it might be possible to screen for problematic data. In this article we focus on leading digit analysis (i.e., Newcomb-Benford Law) and terminal digit analysis to detect potentially problematic data.

#### Newcomb-Benford law

The Newcomb-Benford law [NBL; @10.2307/2369148;@10.2307/984802] states that leading digits do not have an equal probability of occuring under certain conditions. A leading digit is the left-most digit of a numeric value, where a digit is any of the nine natural numbers ($1,2,3,...,9$). The distribution of the leading digit, according to the NBL is
$$
P(d)=log_{10}\frac{1+d}{d}
$$
where $d$ is the natural number of the leading digit and $P(d)$ is the probability of $d$ occurring. Table `r tabnr` indicates the expected leading digit distribution based on the NBL. This expected distribution is typically compared to the observed distribution with a $\chi^2$-test ($df=9-1$), which requires a minimum of 45 observations based on the rule of thumb outlined by @agresti2003categorical ($n=I\times J \times 5$, with $I$ rows and $J$ columns). The NBL has been applied to detect financial fraud [e.g., @10.2307/27643897], voting fraud [e.g., @durtschi2004effective], and also to detect problems in scientific data [e.g., @10.1007/s00101-017-0333-1].

<!-- In science, for example, the NBL has been tested on 20 falsified anesthesia papers, detecting 18 as problematic [@10.1007/s00101-012-2029-x], or to test coefficients reported in economics journals [@10.1111/j.1468-0475.2009.00475.x].  -->
```{r, echo=FALSE}
tab <- data.frame(digit = 1:9, prop = log10((1+1:9)/1:9))
names(tab) <- c('Digit', 'Proportion')
knitr::kable(tab, digits = 3, caption = 'The expected first digit distribution, based on the Newcomb-Benford Law.')
tabnr <- tabnr + 1
```

However, given that the NBL only applies under specific conditions that are rarely fulfilled in the social sciences, its applicability for detecting data anomalies in science can be questioned. First, the NBL only applies for true ratio scale measures [@10.2307/2246134;@10.1214/11-PS175]. Second, sufficient range on the measure is required for the NBL to apply [i.e., range from $1-1000000$ ;@10.1198/tast.2009.0005]. Third, these measures should not be subject to digit preferences, for example due to psychological preferences for rounded numbers. Fourth, any form of truncation undermines the NBL [@10.1515/9781400866595-011]. Moreover, some research has even indicated humans might be sensitive to fabricating data that are in line with the NBL [@10.1080/02664760601004940;@Burns2009], immediately undermining the applicability of the NBL. Nonetheless, considering the four conditions for the NBL to apply, we preregistered that this method would not prove fruitful [@10.3897/rio.2.e8860].

#### Terminal digit analysis

Terminal digit analysis is based on the principle that the rightmost digit is the most random digit of a number, hence, is expected to be uniformly distributed under specific conditions [@10.1080/08989629508573866;@10.1080/03610919608813325]. Terminal digit analysis is conducted with a $\chi^2$-test ($df=10-1$) on the digit occurrence counts (including zero), where the observed frequencies are compared with the expected uniform frequencies. Tthe rule of thumb outlined by @agresti2003categorical indicates at least 50 observations are required to provide a meaningful test of the terminal digit distribution ($n=I\times J \times 5$, with $I$ rows and $J$ columns). Terminal digit analysis was developed during the Imanishi-Kari case by @10.1080/03610919608813325 [for a history of this decade long case, see @kevles2000baltimore].

As an example, Figure `r fignr` depicts the digit counts for the first- through fifth digit of a random, normally distributed variable. The first- and second digit distributions are clearly non-uniform, whereas the third-, fourth-, and fifth digit distributions are uniformly distributed.

```{r, echo=FALSE, fig.align="center", fig.asp=.5, out.width='100%', fig.cap="Illustration of how digit distributions evolve from first- through later digits. We sampled 100,000 values from a normal distribution, N~(0, 1.5), to create these digit distributions."}
set.seed(1234)
tmp1 <- 0
tmp2 <- 1.5
y <- data.frame(value = rnorm(100000, tmp1, tmp2))
x <- abs(y$value)
splitted <- strsplit(as.character(x), "")
df1 <- data.frame(digit = as.numeric(unlist(lapply(splitted, function(x) return(x[1])))))
df1 <- data.frame(digit = df1[!df1$digit == 0,])
df2 <- data.frame(digit = as.numeric(unlist(lapply(splitted, function(x) return(x[3])))))
df3 <- data.frame(digit = as.numeric(unlist(lapply(splitted, function(x) return(x[4])))))
df4 <- data.frame(digit = as.numeric(unlist(lapply(splitted, function(x) return(x[5])))))
df5 <- data.frame(digit = as.numeric(unlist(lapply(splitted, function(x) return(x[6])))))

overall <- ggplot(y, aes(x = value)) +
  geom_density(alpha = .3, fill = 'red') + 
  xlab(sprintf("Value from N~(%s,%s)", tmp1, tmp2)) +
  ylab("Density") +
  theme(axis.title=element_text(size=10))
